---
title: "QC report: `r params$samplename`"
subtitle: "`r v <- tryCatch({ri <- read.delim(params$runinfo, colClasses = 'character'); ri$value[ri$parameter == 'mihcro version'][1]}, error = function(e) NA); paste0(if (!is.na(v) && nzchar(v)) paste0('mihcro ', v, ' · ') else '', 'rendered ', format(Sys.Date(), '%d %B %Y'))`"
params:
  samplename:
  cellbyfeature:
  markerfile:
  roiclasses: ""
  runinfo: ""
  scalejson: ""
  patches: ""
output:
  html_document:
    theme: flatly
    number_sections: true
    md_extensions: -smart
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(dplyr)
library(ggplot2)
library(tidyr)
library(ggcorrplot)
library(Seurat)
library(tibble)
library(pheatmap)
library(kableExtra)

has_file <- function(path) !is.null(path) && nzchar(path) && file.exists(path)

samplename <- params$samplename

# Run information written by the pipeline: group, parameter, value
run_info <- if (has_file(params$runinfo)) {
  read.delim(params$runinfo, colClasses = "character", na.strings = character(0))
} else {
  data.frame(group = character(0), parameter = character(0), value = character(0))
}
run_param <- function(name, default = "") {
  value <- run_info$value[run_info$parameter == name]
  if (length(value) == 0 || is.na(value[1]) || !nzchar(value[1])) default else value[1]
}

# Pixel size of the processed image (MCQuant areas are in pixels)
scale_info <- if (has_file(params$scalejson)) jsonlite::fromJSON(params$scalejson) else list()
mpp <- if (!is.null(scale_info$final_physical_size_x)) as.numeric(scale_info$final_physical_size_x) else NA_real_
area_factor <- if (is.na(mpp)) 1 else mpp^2
area_unit <- if (is.na(mpp)) "px²" else "µm²"

# Patches segmented
patches <- if (has_file(params$patches)) read.csv(params$patches) else NULL

# Cell-by-feature table
cells <- read.csv(params$cellbyfeature, na.strings = c("", "inf"))
colnames(cells) <- colnames(cells) %>% make.names() %>% trimws()
n_cells_raw <- nrow(cells)
cells <- cells %>% drop_na()
n_removed_na <- n_cells_raw - nrow(cells)

markerfile <- read.csv(params$markerfile, header = TRUE)
markerfile$marker_name <- markerfile$marker_name %>% make.names() %>% trimws()
missing_markers <- setdiff(markerfile$marker_name, colnames(cells))
eff_markers <- intersect(markerfile$marker_name, colnames(cells))
num.markers <- length(eff_markers)

marker_long <- cells %>%
  select(CellID, all_of(eff_markers)) %>%
  pivot_longer(cols = -CellID, names_to = "Marker", values_to = "Value") %>%
  mutate(Marker = factor(Marker, levels = eff_markers))

# Cells with implausible intensities (overflowed quantification) are removed
flagged <- marker_long %>% group_by(CellID) %>% filter(sum(Value) > 1e+30) %>% pull(CellID) %>% unique()
if (length(flagged) > 0) {
  cells <- cells %>% filter(!CellID %in% flagged)
  marker_long <- marker_long %>% filter(!CellID %in% flagged)
}

marker_clr <- marker_long %>%
  group_by(CellID) %>%
  mutate(Value_CLR = log(Value + 1) - mean(log(Value + 1), na.rm = TRUE)) %>%
  ungroup()

cells$Area_scaled <- cells$Area * area_factor

# Regions of interest (QuPath annotations or automatic tissue detection)
has_roi <- "roi_class" %in% colnames(cells)
roi_classes <- NULL
if (has_roi && has_file(params$roiclasses)) {
  roi_all <- read.csv(params$roiclasses)
  roi_total <- roi_all[which(roi_all$index == 0), ]
  roi_classes <- roi_all[which(!as.logical(roi_all$excluded) & roi_all$index > 0), ]
  roi_classes$column <- make.names(roi_classes$column)
}

# Tables: rounded numbers, thousands separators, one style throughout
fmt_table <- function(df, caption = NULL, digits = 3) {
  df %>%
    mutate(across(where(is.double), ~ signif(.x, digits))) %>%
    kbl(caption = caption, format.args = list(big.mark = ",")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
}

# Highlighted notice box
notice <- function(text) cat(sprintf('<div class="alert alert-warning" role="alert">%s</div>\n\n', text))

# Cluster colours: the fixed 30-colour palette first, then generated colours for any extra clusters
base_pal <- c("#e6194b", "#3cb44b", "#ffe119", "#4363d8", "#f58231", "#911eb4", "#42d4f4", "#f032e6", "#bfef45", "#fabed4",
              "#469990", "#dcbeff", "#9a6324", "#fffac8", "#800000", "#aaffc3", "#808000", "#ffd8b1", "#000075", "#a9a9a9",
              "#993F00", "#4C005C", "#005C31", "#2BCE48", "#94FFB5", "#C20088", "#FFA405", "#5EF1F2", "#00998F", "lightgrey")
cluster_palette <- function(levels) {
  levels <- as.character(levels)
  n <- length(levels)
  colours <- if (n <= length(base_pal)) base_pal[seq_len(n)] else c(base_pal, grDevices::hcl.colors(n - length(base_pal), "Dynamic"))
  setNames(colours, levels)
}

# pheatmap opens and closes a scratch device internally, which leaves the outer chunk's device
# current when called from a child document; build the heatmap silently and draw it on the caller's device
draw_pheatmap <- function(...) {
  dev <- grDevices::dev.cur()
  ph <- pheatmap(..., silent = TRUE)
  if (dev != 1) grDevices::dev.set(dev)
  grid::grid.newpage()
  grid::grid.draw(ph$gtable)
  invisible(ph)
}
```

# Overview

## Sample summary

```{r overview, results='asis'}
segmentation <- run_param("segmentation", "not recorded")

roi_source <- if (!has_roi) {
  "None (whole image)"
} else if (!is.null(roi_classes) && identical(roi_classes$class, "Tissue")) {
  "Automatic tissue detection"
} else if (!is.null(roi_classes)) {
  sprintf("QuPath annotations (%d class%s)", nrow(roi_classes), if (nrow(roi_classes) == 1) "" else "es")
} else {
  "Yes"
}

image_size <- if (!is.null(scale_info$output_shape)) {
  dims <- as.numeric(scale_info$output_shape)
  size <- sprintf("%s × %s px", format(dims[3], big.mark = ","), format(dims[2], big.mark = ","))
  if (!is.na(mpp)) size <- sprintf("%s (%.1f × %.1f mm)", size, dims[3] * mpp / 1000, dims[2] * mpp / 1000)
  size
} else {
  "not recorded"
}

summary_rows <- tribble(
  ~Item, ~Value,
  "Sample", samplename,
  "Cells analysed", format(nrow(cells), big.mark = ","),
  "Segmentation", segmentation,
  "Pixel size", if (is.na(mpp)) "not recorded" else sprintf("%.3g µm/px", mpp),
  "Image size", image_size,
  "Markers analysed", sprintf("%d (%s)", num.markers, paste(eff_markers, collapse = ", ")),
  "Region of interest", roi_source
)
if (has_roi && exists("roi_total") && nrow(roi_total) == 1) {
  summary_rows <- add_row(summary_rows, Item = "Region of interest area", Value = sprintf("%.3g mm²", roi_total$area_mm2))
}
if (!is.null(patches)) {
  n_kept <- sum(patches$status == "kept")
  summary_rows <- add_row(summary_rows, Item = "Patches segmented",
                          Value = sprintf("%d of %d (%.0f%%)", n_kept, nrow(patches), 100 * n_kept / nrow(patches)))
}
print(fmt_table(summary_rows))

if (length(missing_markers) > 0) {
  notice(sprintf("<b>Markers not found in the data:</b> %s. They were excluded from the analysis.",
                 paste(missing_markers, collapse = ", ")))
}
if (n_removed_na > 0) {
  notice(sprintf("<b>%s cell%s</b> with missing or infinite values %s removed.", format(n_removed_na, big.mark = ","),
                 if (n_removed_na == 1) "" else "s", if (n_removed_na == 1) "was" else "were"))
}
if (length(flagged) > 0) {
  notice(sprintf("<b>%s cell%s</b> with implausibly high intensities (total above 1e30) %s removed.",
                 format(length(flagged), big.mark = ","), if (length(flagged) == 1) "" else "s",
                 if (length(flagged) == 1) "was" else "were"))
}
```

## Workflow parameters

```{r parameters, results='asis'}
if (nrow(run_info) == 0) {
  cat("Workflow parameters were not recorded for this report.\n\n")
} else {
  bg_method <- run_param("dapi_bg_method", "none")
  bg_needs <- list(
    none = character(0), otsu_only = "dapi_otsu_leniency", mean = "dapi_otsu_leniency",
    gaussian = c("dapi_bg_sigma", "dapi_otsu_leniency"), rollingball = c("dapi_bg_radius", "dapi_otsu_leniency"),
    af = c("af_channel", "dapi_otsu_leniency")
  )
  hidden <- setdiff(c("dapi_bg_sigma", "dapi_bg_radius", "dapi_otsu_leniency", "af_channel"), bg_needs[[bg_method]])
  if (segmentation != "cellpose") hidden <- c(hidden, "cellpose_model", "cellpose_diam")

  shown <- run_info %>%
    filter(!parameter %in% hidden) %>%
    mutate(
      value = case_when(
        parameter == "membrane_channel" & value == "" ~ "none",
        parameter == "cellpose_model" & value == "" ~ "cyto3 (bundled)",
        value == "" ~ "not set",
        TRUE ~ value
      ),
      Parameter = ifelse(grepl(" ", parameter), parameter, paste0("<code>--", parameter, "</code>"))
    )
  groups <- rle(shown$group)
  shown %>%
    select(Parameter, Value = value) %>%
    kbl(escape = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left") %>%
    pack_rows(index = setNames(groups$lengths, groups$values)) %>%
    print()
}
```

```{r roi-summary, results='asis', eval=has_roi}
cat("# Regions of interest {.tabset}\n\n## Summary {-}\n\n")

primary_counts <- cells %>% count(roi_class, name = "Cells (primary region)")

if (!is.null(roi_classes)) {
  roi_summary <- data.frame(
    Region = roi_classes$class,
    `Area (mm²)` = roi_classes$area_mm2,
    `Cells in region` = sapply(roi_classes$column, function(col) {
      if (col %in% colnames(cells)) sum(as.logical(cells[[col]]), na.rm = TRUE) else NA
    }),
    check.names = FALSE
  )
  roi_summary$`Cells per mm²` <- round(roi_summary$`Cells in region` / roi_summary$`Area (mm²)`, 1)
  roi_summary <- roi_summary %>% left_join(primary_counts, by = c("Region" = "roi_class"))
} else {
  roi_summary <- primary_counts %>% rename(Region = roi_class)
}

print(fmt_table(roi_summary, "Cells per region. A cell counts towards every region containing its centroid; its primary region is the smallest one."))
```

```{r roi-spatial, results='asis', eval=has_roi, fig.dim=c(10,10)}
cat("\n\n## Spatial distribution {-}\n\n")

roi_plot_df <- cells %>% select(X_centroid, Y_centroid, roi_class)
if (nrow(roi_plot_df) > 200000) {
  roi_plot_df <- roi_plot_df[sample(nrow(roi_plot_df), 200000), ]
}

print(
  ggplot(roi_plot_df, aes(x = X_centroid, y = Y_centroid, color = roi_class)) +
    geom_point(size = 0.2) +
    coord_fixed() +
    scale_y_reverse() +
    theme_minimal() +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    labs(title = "Cell centroids by primary region", x = "X (px)", y = "Y (px)", color = "Region")
)
```

```{r roi-markers, results='asis', eval=has_roi, fig.dim=c(12,8)}
cat("\n\n## Marker intensity by region {-}\n\n")

roi_lookup <- data.frame(CellID = cells$CellID, roi_class = cells$roi_class)
clr_roi <- marker_clr %>% inner_join(roi_lookup, by = "CellID")

print(
  ggplot(clr_roi, aes(x = roi_class, y = Value_CLR, fill = roi_class)) +
    geom_boxplot(outlier.size = 0.2) +
    facet_wrap(~Marker, scales = "free_y", ncol = 4) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "CLR-transformed marker intensity by primary region", x = NULL, y = "intensity", fill = "Region")
)
```

# Morphology {.tabset}

## Cell area {-}

```{r}
cells %>%
  summarise(
    Cells = n(),
    Mean = mean(Area_scaled),
    Median = median(Area_scaled),
    SD = sd(Area_scaled),
    Min = min(Area_scaled),
    Max = max(Area_scaled)
  ) %>%
  fmt_table(sprintf("Cell area (%s)", area_unit))
```

```{r}
ggplot(cells, aes(x = Area_scaled)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Area distribution", x = sprintf("Area (%s)", area_unit), y = "Density")

ggplot(cells, aes(x = Area_scaled)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  scale_x_log10() +
  labs(title = "Area distribution (log scale)", x = sprintf("Area (%s, log scale)", area_unit), y = "Density")
```

## Eccentricity {-}

```{r}
ggplot(cells, aes(x = Eccentricity)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Eccentricity distribution", x = "Eccentricity", y = "Density")
```

## Solidity {-}

```{r}
ggplot(cells, aes(x = Solidity)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Solidity distribution", x = "Solidity", y = "Density")
```

## Extent {-}

```{r}
ggplot(cells, aes(x = Extent)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Extent distribution", x = "Extent", y = "Density")
```

# Marker intensity statistics {.tabset}

## Raw {-}

```{r}
marker_long %>%
  group_by(Marker) %>%
  summarise(Mean = mean(Value), Median = median(Value), SD = sd(Value), Min = min(Value), Max = max(Value)) %>%
  fmt_table("Channel intensity (raw)")
```

## CLR {-}

```{r}
marker_clr %>%
  group_by(Marker) %>%
  summarise(Mean = mean(Value_CLR), Median = median(Value_CLR), SD = sd(Value_CLR), Min = min(Value_CLR), Max = max(Value_CLR)) %>%
  fmt_table("Channel intensity (CLR-transformed)")
```

# Marker intensity distributions {.tabset}

```{r}
# 4 facets per row, 5/3 in per row of facets
eff.fig.width <- 10
eff.fig.height <- (5 / 3) * ceiling(num.markers / 4)
```

## Raw {-}

```{r, fig.width=eff.fig.width, fig.height=eff.fig.height}
ggplot(marker_long, aes(x = Value, fill = Marker)) +
  geom_density() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "Marker intensity", x = "intensity", y = "density") +
  NoLegend()
```

## CLR {-}

```{r, fig.width=eff.fig.width, fig.height=eff.fig.height}
ggplot(marker_clr, aes(x = Value_CLR, fill = Marker)) +
  geom_density() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "Marker intensity (CLR)", x = "intensity", y = "density") +
  NoLegend()
```

# Marker intensity boxplots {.tabset}

```{r}
eff.fig.width <- max(5, (7 / 10) * num.markers)
eff.fig.height <- 5
```

## Raw {-}

```{r, fig.width=eff.fig.width, fig.height=eff.fig.height}
ggplot(marker_long, aes(x = Marker, y = Value, fill = Marker)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Intensity distribution", x = "markers", y = "intensity") +
  NoLegend() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

## CLR {-}

```{r, fig.width=eff.fig.width, fig.height=eff.fig.height}
ggplot(marker_clr, aes(x = Marker, y = Value_CLR, fill = Marker)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Intensity distribution (CLR)", x = "markers", y = "intensity") +
  NoLegend() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

# Correlation between markers

```{r}
eff.fig.width <- max(5, (7 / 10) * num.markers)
eff.fig.height <- max(4, (5 / 10) * num.markers)
```

```{r, fig.width=eff.fig.width, fig.height=eff.fig.height}
corr_matrix <- cells %>% select(all_of(eff_markers)) %>% cor() %>% round(digits = 2)

ggcorrplot(corr_matrix, hc.order = FALSE, type = "lower", lab = TRUE) +
  ggtitle("Pearson correlation of marker intensities")
```

# QQ plots {.tabset}

```{r}
eff.fig.width <- 7
eff.fig.height <- (5 / 3) * ceiling(num.markers / 4)

# Downsample for plotting only; the full data is used everywhere else
qq_cells <- unique(marker_long$CellID)
if (length(qq_cells) > 100000) qq_cells <- sample(qq_cells, size = 100000, replace = FALSE)
qq_long <- marker_long %>% filter(CellID %in% qq_cells)
qq_clr <- marker_clr %>% filter(CellID %in% qq_cells)
```

## Raw {-}

```{r, fig.width=eff.fig.width, fig.height=eff.fig.height}
ggplot(qq_long, aes(sample = Value, color = Marker)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "QQ plot", x = "Theoretical quantiles", y = "Sample quantiles") +
  NoLegend()
```

## CLR {-}

```{r, fig.width=eff.fig.width, fig.height=eff.fig.height}
ggplot(qq_clr, aes(sample = Value_CLR, color = Marker)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "QQ plot (CLR)", x = "Theoretical quantiles", y = "Sample quantiles") +
  NoLegend()
```

```{r clustering-compute, include=FALSE}
# Seurat object from the CLR-normalised markers; the nuclear channel is not used for clustering
nuclear_marker <- make.names(run_param("nuclear_channel", "DAPI"))
cells$CellID <- make.unique(as.character(cells$CellID))
marker_cols <- setdiff(eff_markers, nuclear_marker)
meta_cols <- setdiff(colnames(cells), marker_cols)

expr <- t(as.matrix(cells[, marker_cols, drop = FALSE]))
colnames(expr) <- cells$CellID
expr[is.na(expr)] <- 0

meta_df <- cells[, meta_cols, drop = FALSE] %>% column_to_rownames("CellID")

seu <- CreateSeuratObject(counts = expr, assay = "PROT", meta.data = meta_df)
DefaultAssay(seu) <- "PROT"
seu <- NormalizeData(seu, normalization.method = "CLR", margin = 2, verbose = FALSE)
VariableFeatures(seu) <- rownames(seu)
seu <- ScaleData(seu, verbose = FALSE)

use_sketch <- ncol(seu) > 100000
if (use_sketch) {
  # Cluster a 50,000-cell sketch, then project the clusters onto all cells
  seu <- SketchData(seu, ncells = 50000, method = "LeverageScore", sketched.assay = "sketch")
  DefaultAssay(seu) <- "sketch"
  seu <- ScaleData(seu, verbose = FALSE)
  seu <- RunPCA(seu, npcs = 6, verbose = FALSE)
  pcs_use <- 1:min(6, ncol(Embeddings(seu, "pca")))
  seu <- RunUMAP(seu, dims = pcs_use, verbose = FALSE, reduction.name = "umap.sketch", reduction.key = "UMAP_sketch_")
  seu <- FindNeighbors(seu, dims = pcs_use, verbose = FALSE)
  seu <- FindClusters(seu, resolution = c(0.25, 0.5, 0.75, 1), verbose = FALSE)
  seu <- ProjectData(
    object = seu,
    assay = "PROT",
    full.reduction = "pca.full",
    sketched.assay = "sketch",
    sketched.reduction = "pca",
    umap.model = "umap.sketch",
    dims = pcs_use,
    refdata = list(
      PROT_snn_res.0.25 = "sketch_snn_res.0.25",
      PROT_snn_res.0.5 = "sketch_snn_res.0.5",
      PROT_snn_res.0.75 = "sketch_snn_res.0.75",
      PROT_snn_res.1 = "sketch_snn_res.1"
    )
  )
  DefaultAssay(seu) <- "PROT"
} else {
  seu <- RunPCA(seu, npcs = 6, features = VariableFeatures(seu), verbose = FALSE)
  pcs_use <- 1:min(6, ncol(Embeddings(seu, "pca")))
  seu <- FindNeighbors(seu, reduction = "pca", dims = pcs_use, verbose = FALSE)
  seu <- RunUMAP(seu, dims = pcs_use, verbose = FALSE)
  seu <- FindClusters(seu, resolution = c(0.25, 0.5, 0.75, 1), verbose = FALSE)
}
umap_red <- if (use_sketch) "umap.sketch" else "umap"

# Cell positions for the spatial plots
centroids <- seu[[]] %>%
  rownames_to_column("bc") %>%
  mutate(y = X_centroid, x = Y_centroid, cell = bc) %>%
  dplyr::select(x, y, cell)
seu[["myfov"]] <- CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")

# Saved for downstream analysis
saveRDS(seu, file = paste0(samplename, "_seurat.rds"))
```

# Spatial marker expression

```{r, fig.dim=c(10,10)}
ImageFeaturePlot(seu, features = marker_cols, combine = TRUE) &
  scale_y_reverse() &
  theme(plot.background = element_rect(fill = "black"))
```

# Clustering (CLR-transformed data) {.tabset .tabset-pills}

```{r clustering, results='asis'}
for (resolution in c(0.25, 0.5, 0.75, 1)) {
  cat(knitr::knit_child("QCreport_resolution.Rmd", envir = environment(), quiet = TRUE), sep = "\n")
}
```
