---
title: "QC report: `r params$samplename`"
params:
  samplename:
  cellbyfeature:
  markerfile:
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    code_folding: hide
---

# setup

```{r, warning=FALSE, include=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(ggcorrplot)
library(Seurat)
library(tibble)
library(pheatmap)
library(kableExtra)

samplename <- params$samplename
message("Running QC Report for:", samplename)

cbf_path <- params$cellbyfeature
message("Cell by feature csv at:", cbf_path)
mesmer_data <- read.csv(cbf_path, na.strings=c("", "inf"))
colnames(mesmer_data) <- make.names(colnames(mesmer_data))
mesmer_data <- mesmer_data %>% drop_na()

mark_path <- params$markerfile
message("Markerfile at:", mark_path)
markerfile <- read.csv(mark_path, header=TRUE)
marker_number <- nrow(markerfile)
markerfile$marker_name <- make.names(markerfile$marker_name)


mesmer_marker <- mesmer_data %>% select(CellID, all_of(markerfile$marker_name)) %>%
  pivot_longer(cols = -CellID,
               names_to = "Marker",
               values_to = "Value")
```

```{r, echo=FALSE}
flag <- mesmer_marker %>% group_by(CellID) %>% filter(sum(Value)>1e+30) %>% pull(CellID) %>% unique()

if (length(flag)>0) {
  warning("Certain 'cells' appear to have unrealistically high intensities and will be removed.\nProblematic rows:")
  mesmer_data %>% filter(CellID %in% flag)
  mesmer_data <- mesmer_data %>% filter(!CellID %in% flag)
  mesmer_marker <- mesmer_marker %>% filter(!CellID %in% flag)
}
```

# descriptive statistics {.tabset}
## raw
```{r, echo=FALSE}
mesmer_summary <- mesmer_marker %>%
  group_by(Marker) %>%
  summarise(
    Mean = mean(Value),
    Median = median(Value),
    SD = sd(Value),
    Min = min(Value),
    Max = max(Value)
  )
mesmer_summary %>%
  kbl(caption = "Statistical Table") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

## log transformation
```{r, echo=FALSE}
mesmer_log_summary <- mesmer_marker %>% mutate(Value=log1p(Value)) %>%
  group_by(Marker) %>%
  summarise(
    Mean = mean(Value),
    Median = median(Value),
    SD = sd(Value),
    Min = min(Value),
    Max = max(Value)
  )
mesmer_log_summary %>%
  kbl(caption = "Statistical Table") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

## CLR transformation
```{r, echo=FALSE}
df_clr_mesmer <- mesmer_marker %>%
  group_by(CellID) %>%
  mutate(Value_CLR = log(Value + 1) - mean(log(Value + 1), na.rm = TRUE)) %>%
  ungroup()

mesmer_log_summary <- df_clr_mesmer %>%
  group_by(Marker) %>%
  summarise(
    Mean = mean(Value_CLR),
    Median = median(Value_CLR),
    SD = sd(Value_CLR),
    Min = min(Value_CLR),
    Max = max(Value_CLR)
  )
mesmer_log_summary %>%
  kbl(caption = "Statistical Table") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

# Histogram of marker intensity {.tabset}
## raw
```{r, echo=FALSE}
ggplot(mesmer_marker, aes(x = Value, fill = Marker)) +
  geom_density() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "Marker intensity",
       x = "intensity", y = "density")
```

## log transformation
```{r, echo=FALSE}
ggplot(mesmer_marker %>% mutate(Value=log1p(Value)), aes(x = Value, fill = Marker)) +
  geom_density() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "Marker intensity",
       x = "intensity", y = "density")
```

## CLR transformation
```{r, echo=FALSE}
df_clr_mesmer <- mesmer_marker %>%
  group_by(CellID) %>%
  mutate(Value_CLR = log(Value + 1) - mean(log(Value + 1), na.rm = TRUE)) %>%
  ungroup()

ggplot(df_clr_mesmer, aes(x = Value_CLR, fill = Marker)) +
  geom_density() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "Marker intensity",
       x = "intensity", y = "density")
```

# Boxplot {.tabset}
## raw
```{r, echo=FALSE}
ggplot(mesmer_marker, aes(x = Marker, y = Value, fill = Marker)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of intensity distribution",
       x = "markers", y = "intensity")
```

## log transformation
```{r, echo=FALSE}
ggplot(mesmer_marker %>% mutate(Value=log1p(Value)), aes(x = Marker, y = Value, fill = Marker)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of intensity distribution",
       x = "markers", y = "intensity")
```

## CLR transformation
```{r, echo=FALSE}
df_clr_mesmer <- mesmer_marker %>%
  group_by(CellID) %>%
  mutate(Value_CLR = log(Value + 1) - mean(log(Value + 1), na.rm = TRUE)) %>%
  ungroup()

ggplot(df_clr_mesmer, aes(x = Marker, y = Value_CLR, fill = Marker)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of intensity distribution",
       x = "markers", y = "intensity")
```

# correlation between markers {.tabset}

```{r, echo=FALSE}

mesmer_corr_matrix <- mesmer_data %>% select(all_of(markerfile$marker_name)) %>%
  cor() %>%
  round(digits=2)

ggcorrplot(mesmer_corr_matrix, hc.order =FALSE,
           type ="lower", lab =TRUE)+
  ggtitle("Pearson Correlation Heatmap of Marker Intensities")
```

# median marker intensity heatmap
```{r, echo=FALSE}
mesmer_median <- select(mesmer_summary, Marker, Median) %>% mutate(Sample = samplename)
ggplot(mesmer_median, aes(x = Sample, y = Marker, fill = Median, scale = "row")) +
  geom_tile(color = "white") +
  scale_fill_gradientn(colors = c("steelblue", "white", "indianred1")) +
  theme_minimal() +
  labs(title = "median marker intensity heatmap",
       x = "Sample", y = "Marker")

```

# QQ plot {.tabset}
## raw
```{r, echo=FALSE}
ggplot(mesmer_marker, aes(sample = Value, fill = Marker, color = Marker)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "QQ plot",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

## log transformation
```{r, echo=FALSE}
ggplot(mesmer_marker %>% mutate(Value=log1p(Value)), aes(sample = Value, color = Marker)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "QQ plot",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

## arcsin transformation
```{r, echo=FALSE}
asinh.transform = function(x, cofactor) {
  return(asinh(x/cofactor)) }
ggplot(mesmer_marker %>% mutate(Value=asinh.transform(Value, 200)), aes(sample = Value, color = Marker)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "QQ plot",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

## CLR transformation
```{r, echo=FALSE}
df_clr_mesmer <- mesmer_marker %>%
  group_by(CellID) %>%
  mutate(Value_CLR = log(Value + 1) - mean(log(Value + 1), na.rm = TRUE)) %>%
  ungroup()

ggplot(df_clr_mesmer, aes(sample = Value_CLR, color = Marker)) +
  stat_qq() +
  stat_qq_line() +
  theme_minimal() +
  facet_wrap(~Marker, scales = "free", ncol = 4) +
  labs(title = "QQ plot",
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```


# Clustering analysis (CLR-transformed data) {.tabset .tabset-pills}

```{r, echo=FALSE, warning=FALSE, include=F}
mesmer_data$CellID <- make.unique(as.character(mesmer_data$CellID)) # make sure the cell ID is unique
marker_cols <- markerfile$marker_name
meta_cols <- setdiff(colnames(mesmer_data), marker_cols)

# build marker matrix
expr_mesmer <- t(as.matrix(mesmer_data[, marker_cols, drop = FALSE]))
colnames(expr_mesmer) <- mesmer_data$CellID
expr_mesmer[is.na(expr_mesmer)] <- 0

# build meta.data
meta_df_mesmer <- mesmer_data[, meta_cols, drop = FALSE] %>% column_to_rownames("CellID")

# create Seurat object
seu_mesmer <- CreateSeuratObject(
  counts    = expr_mesmer,
  assay     = "PROT", # name the assay as PROT
  meta.data = meta_df_mesmer
)

# normalization and feature selection
DefaultAssay(seu_mesmer) <- "PROT"
seu_mesmer <- NormalizeData(seu_mesmer, normalization.method = "CLR", margin = 2, verbose = FALSE) # CLR transformation
VariableFeatures(seu_mesmer) <- rownames(seu_mesmer) # choose all marker as targeted feature
seu_mesmer <- ScaleData(seu_mesmer, verbose = FALSE) # z-score (x - mean)/sd

# dimentionality reduction and clustering
seu_mesmer <- RunPCA(seu_mesmer, npcs = 6, features = VariableFeatures(seu_mesmer))
pcs_use <- 1:min(6, ncol(Embeddings(seu_mesmer, "pca"))) # choose the number of PC
seu_mesmer <- FindNeighbors(seu_mesmer, reduction = "pca", dims = pcs_use)
seu_mesmer <- RunUMAP(seu_mesmer, dims = pcs_use)
seu_mesmer <- FindClusters(seu_mesmer, resolution = seq(from=0.2, to=0.8, by=0.1), verbose = FALSE)
```

## resolution = 0.2

```{r, echo = FALSE, warning=FALSE, fig.dim=c(10,10)}
resolution=0.2
seu_mesmer$seurat_clusters <- seu_mesmer[[paste0("PROT_snn_res.", resolution)]]
Idents(seu_mesmer) <- "seurat_clusters"
# visualization
## umap
DimPlot(seu_mesmer, reduction = "umap", label = TRUE, label.box = TRUE)
## tissue image
centroids = seu_mesmer[[]] %>% rownames_to_column("bc") %>% mutate(y=X_centroid, x=Y_centroid, cell=bc) %>% dplyr::select(x,y,cell)
ff=CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")
seu_mesmer[["myfov"]]=ff
ImageDimPlot(seu_mesmer, size=1) + scale_y_reverse()
ImageFeaturePlot(seu_mesmer, features = marker_cols, combine = TRUE) & scale_y_reverse() & theme(plot.background = element_rect(fill="black"))

#df_plot <- FetchData(seu_mesmer, vars = c("X_centroid","Y_centroid","seurat_clusters")) # extract variables
#df_plot=seu_mesmer[[]]
#ggplot(df_plot, aes(x = X_centroid, y = Y_centroid, color = res_0.2)) +
#  geom_point(size = 0.6) +
#  coord_fixed() + # coordinate ratio: 1:1
#  scale_y_reverse() +
#  theme_minimal()

## heatmap
avg <- AverageExpression(
  seu_mesmer,
  assays = "PROT",
  group.by = "seurat_clusters",
  slot = "data"
)$PROT
pheatmap(avg, cluster_rows = FALSE, cluster_cols = FALSE, labels_col = levels(Idents(seu_mesmer)), scale = "row")

## violet plot

VlnPlot(
  seu_mesmer,
  features = marker_cols,
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3,
  same.y.lims = TRUE
)
```

## resolution = 0.3

```{r, echo = FALSE, warning=FALSE, fig.dim=c(10,10)}
resolution=0.3
seu_mesmer$seurat_clusters <- seu_mesmer[[paste0("PROT_snn_res.", resolution)]]
Idents(seu_mesmer) <- "seurat_clusters"
# visualization
## umap
DimPlot(seu_mesmer, reduction = "umap", label = TRUE, label.box = TRUE)
## tissue image
centroids = seu_mesmer[[]] %>% rownames_to_column("bc") %>% mutate(y=X_centroid, x=Y_centroid, cell=bc) %>% dplyr::select(x,y,cell)
ff=CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")
seu_mesmer[["myfov"]]=ff
ImageDimPlot(seu_mesmer, size=1) + scale_y_reverse()
ImageFeaturePlot(seu_mesmer, features = marker_cols, combine = TRUE) & scale_y_reverse() & theme(plot.background = element_rect(fill="black"))

#df_plot <- FetchData(seu_mesmer, vars = c("X_centroid","Y_centroid","seurat_clusters")) # extract variables
#df_plot=seu_mesmer[[]]
#ggplot(df_plot, aes(x = X_centroid, y = Y_centroid, color = res_0.2)) +
#  geom_point(size = 0.6) +
#  coord_fixed() + # coordinate ratio: 1:1
#  scale_y_reverse() +
#  theme_minimal()

## heatmap
avg <- AverageExpression(
  seu_mesmer,
  assays = "PROT",
  group.by = "seurat_clusters",
  slot = "data"
)$PROT
pheatmap(avg, cluster_rows = FALSE, cluster_cols = FALSE, labels_col = levels(Idents(seu_mesmer)), scale = "row")

## violet plot

VlnPlot(
  seu_mesmer,
  features = marker_cols,
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3,
  same.y.lims = TRUE
)
```


## resolution = 0.4

```{r, echo = FALSE, warning=FALSE, fig.dim=c(10,10)}
resolution=0.4
seu_mesmer$seurat_clusters <- seu_mesmer[[paste0("PROT_snn_res.", resolution)]]
Idents(seu_mesmer) <- "seurat_clusters"
# visualization
## umap
DimPlot(seu_mesmer, reduction = "umap", label = TRUE, label.box = TRUE)
## tissue image
centroids = seu_mesmer[[]] %>% rownames_to_column("bc") %>% mutate(y=X_centroid, x=Y_centroid, cell=bc) %>% dplyr::select(x,y,cell)
ff=CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")
seu_mesmer[["myfov"]]=ff
ImageDimPlot(seu_mesmer, size=1) + scale_y_reverse()
ImageFeaturePlot(seu_mesmer, features = marker_cols, combine = TRUE) & scale_y_reverse() & theme(plot.background = element_rect(fill="black"))

#df_plot <- FetchData(seu_mesmer, vars = c("X_centroid","Y_centroid","seurat_clusters")) # extract variables
#df_plot=seu_mesmer[[]]
#ggplot(df_plot, aes(x = X_centroid, y = Y_centroid, color = res_0.2)) +
#  geom_point(size = 0.6) +
#  coord_fixed() + # coordinate ratio: 1:1
#  scale_y_reverse() +
#  theme_minimal()

## heatmap
avg <- AverageExpression(
  seu_mesmer,
  assays = "PROT",
  group.by = "seurat_clusters",
  slot = "data"
)$PROT
pheatmap(avg, cluster_rows = FALSE, cluster_cols = FALSE, labels_col = levels(Idents(seu_mesmer)), scale = "row")

## violet plot

VlnPlot(
  seu_mesmer,
  features = marker_cols,
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3,
  same.y.lims = TRUE
)
```


## resolution = 0.5

```{r, echo = FALSE, warning=FALSE, fig.dim=c(10,10)}
resolution=0.5
seu_mesmer$seurat_clusters <- seu_mesmer[[paste0("PROT_snn_res.", resolution)]]
Idents(seu_mesmer) <- "seurat_clusters"
# visualization
## umap
DimPlot(seu_mesmer, reduction = "umap", label = TRUE, label.box = TRUE)
## tissue image
centroids = seu_mesmer[[]] %>% rownames_to_column("bc") %>% mutate(y=X_centroid, x=Y_centroid, cell=bc) %>% dplyr::select(x,y,cell)
ff=CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")
seu_mesmer[["myfov"]]=ff
ImageDimPlot(seu_mesmer, size=1) + scale_y_reverse()
ImageFeaturePlot(seu_mesmer, features = marker_cols, combine = TRUE) & scale_y_reverse() & theme(plot.background = element_rect(fill="black"))

#df_plot <- FetchData(seu_mesmer, vars = c("X_centroid","Y_centroid","seurat_clusters")) # extract variables
#df_plot=seu_mesmer[[]]
#ggplot(df_plot, aes(x = X_centroid, y = Y_centroid, color = res_0.2)) +
#  geom_point(size = 0.6) +
#  coord_fixed() + # coordinate ratio: 1:1
#  scale_y_reverse() +
#  theme_minimal()

## heatmap
avg <- AverageExpression(
  seu_mesmer,
  assays = "PROT",
  group.by = "seurat_clusters",
  slot = "data"
)$PROT
pheatmap(avg, cluster_rows = FALSE, cluster_cols = FALSE, labels_col = levels(Idents(seu_mesmer)), scale = "row")

## violet plot

VlnPlot(
  seu_mesmer,
  features = marker_cols,
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3,
  same.y.lims = TRUE
)
```


## resolution = 0.6

```{r, echo = FALSE, warning=FALSE, fig.dim=c(10,10)}
resolution=0.6
seu_mesmer$seurat_clusters <- seu_mesmer[[paste0("PROT_snn_res.", resolution)]]
Idents(seu_mesmer) <- "seurat_clusters"
# visualization
## umap
DimPlot(seu_mesmer, reduction = "umap", label = TRUE, label.box = TRUE)
## tissue image
centroids = seu_mesmer[[]] %>% rownames_to_column("bc") %>% mutate(y=X_centroid, x=Y_centroid, cell=bc) %>% dplyr::select(x,y,cell)
ff=CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")
seu_mesmer[["myfov"]]=ff
ImageDimPlot(seu_mesmer, size=1) + scale_y_reverse()
ImageFeaturePlot(seu_mesmer, features = marker_cols, combine = TRUE) & scale_y_reverse() & theme(plot.background = element_rect(fill="black"))

#df_plot <- FetchData(seu_mesmer, vars = c("X_centroid","Y_centroid","seurat_clusters")) # extract variables
#df_plot=seu_mesmer[[]]
#ggplot(df_plot, aes(x = X_centroid, y = Y_centroid, color = res_0.2)) +
#  geom_point(size = 0.6) +
#  coord_fixed() + # coordinate ratio: 1:1
#  scale_y_reverse() +
#  theme_minimal()

## heatmap
avg <- AverageExpression(
  seu_mesmer,
  assays = "PROT",
  group.by = "seurat_clusters",
  slot = "data"
)$PROT
pheatmap(avg, cluster_rows = FALSE, cluster_cols = FALSE, labels_col = levels(Idents(seu_mesmer)), scale = "row")

## violet plot

VlnPlot(
  seu_mesmer,
  features = marker_cols,
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3,
  same.y.lims = TRUE
)
```


## resolution = 0.7

```{r, echo = FALSE, warning=FALSE, fig.dim=c(10,10)}
resolution=0.7
seu_mesmer$seurat_clusters <- seu_mesmer[[paste0("PROT_snn_res.", resolution)]]
Idents(seu_mesmer) <- "seurat_clusters"
# visualization
## umap
DimPlot(seu_mesmer, reduction = "umap", label = TRUE, label.box = TRUE)
## tissue image
centroids = seu_mesmer[[]] %>% rownames_to_column("bc") %>% mutate(y=X_centroid, x=Y_centroid, cell=bc) %>% dplyr::select(x,y,cell)
ff=CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")
seu_mesmer[["myfov"]]=ff
ImageDimPlot(seu_mesmer, size=1) + scale_y_reverse()
ImageFeaturePlot(seu_mesmer, features = marker_cols, combine = TRUE) & scale_y_reverse() & theme(plot.background = element_rect(fill="black"))

#df_plot <- FetchData(seu_mesmer, vars = c("X_centroid","Y_centroid","seurat_clusters")) # extract variables
#df_plot=seu_mesmer[[]]
#ggplot(df_plot, aes(x = X_centroid, y = Y_centroid, color = res_0.2)) +
#  geom_point(size = 0.6) +
#  coord_fixed() + # coordinate ratio: 1:1
#  scale_y_reverse() +
#  theme_minimal()

## heatmap
avg <- AverageExpression(
  seu_mesmer,
  assays = "PROT",
  group.by = "seurat_clusters",
  slot = "data"
)$PROT
pheatmap(avg, cluster_rows = FALSE, cluster_cols = FALSE, labels_col = levels(Idents(seu_mesmer)), scale = "row")

## violet plot

VlnPlot(
  seu_mesmer,
  features = marker_cols,
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3,
  same.y.lims = TRUE
)
```


## resolution = 0.8


```{r, echo = FALSE, warning=FALSE, fig.dim=c(10,10)}
resolution=0.8
seu_mesmer$seurat_clusters <- seu_mesmer[[paste0("PROT_snn_res.", resolution)]]
Idents(seu_mesmer) <- "seurat_clusters"
# visualization
## umap
DimPlot(seu_mesmer, reduction = "umap", label = TRUE, label.box = TRUE)
## tissue image
centroids = seu_mesmer[[]] %>% rownames_to_column("bc") %>% mutate(y=X_centroid, x=Y_centroid, cell=bc) %>% dplyr::select(x,y,cell)
ff=CreateFOV(coords = centroids, type = "centroids", key = "fov", assay = "PROT")
seu_mesmer[["myfov"]]=ff
ImageDimPlot(seu_mesmer, size=1) + scale_y_reverse()
ImageFeaturePlot(seu_mesmer, features = marker_cols, combine = TRUE) & scale_y_reverse() & theme(plot.background = element_rect(fill="black"))

# df_plot <- FetchData(seu_mesmer, vars = c("X_centroid","Y_centroid","seurat_clusters")) # extract variables
# df_plot=seu_mesmer[[]]
# ggplot(df_plot, aes(x = X_centroid, y = Y_centroid, color = seurat_clusters)) +
#  geom_point(size = 0.6) +
#  coord_fixed() + # coordinate ratio: 1:1
#  scale_y_reverse() +
#  theme_minimal()

## heatmap
avg <- AverageExpression(
  seu_mesmer,
  assays = "PROT",
  group.by = "seurat_clusters",
  slot = "data"
)$PROT
pheatmap(avg, cluster_rows = FALSE, cluster_cols = FALSE, labels_col = levels(Idents(seu_mesmer)), scale = "row")

## violet plot

VlnPlot(
  seu_mesmer,
  features = marker_cols,
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3,
  same.y.lims = TRUE
)
```


